---
title: "Meeting Notes"
author: "Martin Schobben & Ilja Kocken"
date: "February 27, 2019"
output: pdf_document
---



```{r setup, include=FALSE}



library(Cairo) # package needed to render per mill sign in rmarkdown pdf ouput with option dev='cairoPDF'
library(swirl)


knitr::opts_chunk$set(warning=FALSE, message=FALSE, dev='CairoPDF')

```


## Using base plot and saving graphics device

The first step involves opening a device, creating the canvas to draw the plot, with the call `png()` `pdf()` or `svg()` as the different devices. Whereafter the call `plot()` can be used to plot `r "R"` objects on the previously opened graphics device. 

## Using ggplot and saving with ggsave

The benefits of ggplot are that dimensions and positions can be better fixed. The ggplot is build layer by layer with a dedicated call for the data sources (as a dataframe) and variable selection (`ggplot`), the visual representation of data (`geom_point(), geom_line() ....`) and formatting of the former layer (`theme()`), and, so give a good overview of the different layers added to a plot. After finalising the plot, you can save it with the call `ggsave`.

### Example

For the example we use the following two packages.

```{r }

library(ggplot2)
library(dplyr)

```

We create a dummy y--x data frame with a label for subsetting. By mutating the label with `bquote`, we can introduce mathematical expressions in the facet and legend labels later on.

```{r}
df <- data.frame(x=c(1:10),y=c(1:10), label=c(rep("A",3),rep("B",7)))

df <- df %>% mutate(label=factor(label, levels = c("A","B"), c(bquote(delta^13*C[carb]~"(\u2030 VPDB)"),bquote(delta^18*O[carb]~"(\u2030 VPDB)"))))


```

Here, we create a ggplot with variables as points. We include `df$label` in the aesthetics (`aes`) to subset the variable according to the label by creating different colors (`colour`) for the datapoints.

```{r}
p <- ggplot(df, aes(x,y, colour = label))+
      geom_point()

p
    
```

We can make multiple plot panels based on the label (`df$label`) variable in the dataframe `df`. By setting argument `labeller` to `label_parsed`, we generate the previous defined mathematical expressions.  So, let's say when you want to plot stable carbon and oxygen isotope data in delta notation as two different panels.

```{r}

p <-  p  +  facet_grid(cols= vars(label), scales = "free", space = "free", labeller=label_parsed)
       
p

```

It is also possible to include mathematical expression in the axis title by using `expression`.

```{r}

p <-  p  + xlab(expression(delta^{13}*C[carb]~"(\u2030 VPDB)"))

p
```

Parsing the same expression for the legend text seems more tricky and involves changing two arguments of `scale_colour` and the additional package `scales`.

```{r}
library(scales)

p <-  p  +    scale_colour_discrete(guide = guide_legend(parse = TRUE), labels = parse_format())


p
```

    
      


## Some links discussed during the meeting

Data visualization course: <https://www.datacamp.com/courses/data-visualization-with-ggplot2-1?tap_a=5644-dce66f&tap_s=93618-a68c98>

Data wrangling course: <https://www.datacamp.com/courses/dplyr-data-manipulation-r-tutorial>

Spatial data in R: <http://mazamascience.com/WorkingWithData/?p=1277>

## Some packages discussed during the meeting

```{r}
lapply(list("swirl","ggplot2", "dplyr"), citation)
```



