---
title: "Meeting notes"
output: github_document
date: '2022-03-04'
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
```

In this meeting / workshop, we explain how to make a figure of combined geochemical and dinocyst assemblage data, against depth. The used data is retrieved from:

Sangiorgi, F., Bijl, P.K., Passchier, S. et al. Southern Ocean warming and Wilkes Land ice sheet retreat during the mid-Miocene. Nat Commun 9, 317 (2018). https://doi.org/10.1038/s41467-017-02609-7


# Load packages
```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(readr)
```
# Load datasets

```{r}
# read data files
TEX <- read_csv("https://raw.githubusercontent.com/uu-code-club/meetings/master/2022-03-10/figure_making/Sangiorgi_TEX.csv")
Dino <- read_csv("https://raw.githubusercontent.com/uu-code-club/meetings/master/2022-03-10/figure_making/Sangiorgi_dino.csv")

```
# Fig 1 - TEX86 versus depth
```{r}
# TEX is loaded as dataset and the depth column is mapped to the x-axis, and TEX to the y-axis
ggplot(data = TEX, mapping = aes(x = `Depth mbsf`, y = `TEX`)) +
  # the first geometry, a simple red point per measurement is added
  geom_point(colour = "red") +
  # in a similar way, we add a line, and specify its width
  geom_line(linetype = 2) +
  # An artifical "error" of 0.06 TEX86 units has been added and plotted using geom_ribbon. We make it transparent by changing the alpha
  geom_ribbon(
    mapping = aes(ymin = (TEX - 0.06), ymax = (TEX + 0.06)),
    alpha = 0.25, fill = "blue"
  ) +
  # we add the title off the plot and of the axis
  labs(title = "plot_sample_1", x = "Depth (mbsf)", y = TEX[86] ~ "(-)") +
  # we can specify the breaks of the x-axis
  scale_x_continuous(breaks = seq(75, 420, 25)) +
  # and also the limits of the x and y axis
  coord_cartesian(xlim = c(80, 420), ylim = c(0.2, 0.8)) +
  # finally, to get rid of the standard grey background and other settings, we change the theme tot the "classic" preset
  theme_classic()
```

# Fig 2 - TEX86 versus depth (+BIT & annotation)
```{r}
# TEX is loaded as dataset and the depth column is mapped to the x-axis, and TEX to the y-axis
ggplot(data = TEX, mapping = aes(x = `Depth mbsf`, y = `TEX`)) +
  # we add a point geometry, but change the aestetic mapping, so that the color of the point corresponds to the BIT index (another column within the TEX dataset)
  geom_point(aes(color = `BIT`)) +
  # simple line between points
  geom_line() +
  # add labels
  labs(title = "TEX86_sample_1", x = "Depth (mbsf)", y = expression(TEX[86])) +
  # set breaks
  scale_x_continuous(breaks = seq(75, 420, 25)) +
  # set limits
  coord_cartesian(xlim = c(80, 420), ylim = c(0.4, 0.6)) +
  # add the classic theme
  theme_classic() +
  # we can customize the preset looks of the plot by changing element wihtin the theme. In this example, we change the size and color of the text
  theme(
    axis.title = element_text(size = 12, colour = "black"),
    axis.text = element_text(size = 12, colour = "black"),
    axis.ticks = element_line(colour = "black")
  ) +
  # an annotation is added using annotate. This can be both text or shapes
  annotate("text", x = 325, y = 0.58, label = "Peak BIT") +
  annotate("rect", xmin = 300, xmax = 350, ymin = 0.56, ymax = 0.6, alpha = .2)
```

# Fig 3 - Methane index versus depth
```{r}
ggplot(data = TEX) +
  geom_line(mapping = aes(x = `Depth mbsf`, y = `Methan index`), colour = "blue") +
  labs(title = "plot_sample_2", x = "Depth (mbsf)", y = "Methane Index") +
  scale_x_continuous(breaks = seq(75, 420, 25)) +
  coord_cartesian(xlim = c(80, 420)) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 12, colour = "black"),
    axis.ticks = element_line(colour = "black")
  )
```

# Data transformation for area plot

To combine certain species of operculodinium and Impagidinium, the data has to be grouped.
Then, the remainder is calculated.

Note the transformation of wide format (as the data was principally loaded into R) to long format using the "pivot_longer" function. This allows ggplot to plot the species/genera as a variable (fill / color)

Discussing this workflow in detail is beyond the scope of today's workshop, but
we recommend that you read up on [tidy data](https://r4ds.had.co.nz/tidy-data.html),
[pivoting](https://r4ds.had.co.nz/tidy-data.html#pivoting), and many other data
processing topics in the excellent [R for Data Science](https://r4ds.had.co.nz/).

```{r}
# get data in right order for area plot
dino_species <- c("Selenopemphix antarctica", "Nematosphaeropsis labyrinthus")
dino_genera <- c(imp = "Impagidinium tot", opr = "Operculodinium")

tb_area <- Dino %>%
  select(
    depth = "depth (mbsf)",
    any_of(dino_species),
    starts_with(dino_genera[["imp"]]),
    starts_with(dino_genera[["opr"]])
  ) %>%
  rowwise() %>%
  mutate(
    !!dino_genera[["imp"]] := sum(c_across(starts_with(dino_genera[["imp"]]))),
    !!dino_genera[["opr"]] := sum(c_across(starts_with(dino_genera[["opr"]]))),
    .keep = "unused"
  ) %>%
  mutate(remainder = 100 - sum(c_across(-depth))) %>%
  drop_na() %>%
  pivot_longer(-depth, names_to = "taxa", values_to = "percent")

# let's have a look at what this looks like: (the conversion is just for github so it looks better)
head(as.data.frame(tb_area))
```

# Fig  4 - Area plot versus depth
```{r}
ggplot(data = tb_area, mapping = aes(x = depth, y = percent, fill = taxa)) +
  geom_area(position = "stack", orientation = "x") +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(breaks = seq(75, 420, 25), expand = c(0, 0)) +
  labs(
    fill = "",
    y = "%",
    x = "Depth (mbsf)"
  ) +
  coord_cartesian(xlim = c(80, 420), ylim = c(0, 100)) +
  theme_classic() +
  scale_fill_brewer(palette = "Spectral") +
  theme(
    axis.title = element_text(size = 12, colour = "black"),
    axis.text = element_text(size = 12, colour = "black"),
    axis.ticks = element_line(colour = "black"),
    legend.position = "bottom"
  )
```

# Fig 5 - Hor. combined plots 1,3,4
We first store ggplots as variables (P1,P2,P#), which we then simply combine by P1+P2+P3 using the "patchwork" package.
The figure is saved as an pdf using "ggsave".
```{r}
P1 <- ggplot(data = TEX, mapping = aes(x = `Depth mbsf`, y = `TEX`)) +
  geom_point(colour = "red") +
  geom_line(linetype = 2) +
  geom_ribbon(mapping = aes(ymin = (TEX - 0.06), ymax = (TEX + 0.06)), width = 10, alpha = 0.25, fill = "blue") +
  labs(x = "Depth (mbsf)", y = expression(TEX[86])) +
  scale_x_continuous(position = "top", breaks = seq(75, 420, 25)) +
  coord_cartesian(xlim = c(80, 420), ylim = c(0, 0.8)) +
  theme_classic() +
  theme(
    axis.title = element_text(size = 12, colour = "black"),
    axis.text = element_text(size = 12, colour = "black"),
    axis.ticks = element_line(colour = "black")
  )

P2 <- ggplot(data = TEX) +
  geom_line(mapping = aes(x = `Depth mbsf`, y = `Methan index`), colour = "blue") +
  labs(x = "Depth (mbsf)", y = "Methane Index") +
  scale_x_continuous(breaks = seq(75, 420, 25)) +
  coord_cartesian(xlim = c(80, 420)) +
  theme_classic() +
  theme(
    axis.title.y = element_text(size = 12, colour = "black"),
    axis.text.y = element_text(size = 12, colour = "black"),
    axis.ticks.y = element_line(colour = "black"),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    axis.title.x = element_blank()
  )

P3 <- ggplot(data = tb_area, mapping = aes(x = depth, y = percent, fill = taxa)) +
  geom_area(position = "stack", orientation = "x") +
  geom_segment(alpha = 0.1, aes(y = 100, xend = depth, yend = 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(breaks = seq(75, 420, 25), expand = c(0, 0)) +
  coord_cartesian(xlim = c(80, 420), ylim = c(0, 100)) +
  labs(
    fill = "",
    y = "%",
    x = "Depth (mbsf)"
  ) +
  theme_classic() +
  scale_fill_brewer(palette = "Spectral") +
  theme(
    axis.title = element_text(size = 12, colour = "black"),
    axis.text = element_text(size = 12, colour = "black"),
    axis.ticks = element_line(colour = "black"),
    legend.position = "bottom"
  )

P1 + P2 + P3 + plot_layout(nrow = 3)

ggsave("finalfigh.pdf")
```

# Fig 6 - Vert. combined plots 1,3,4
As for Fig  5, we first store ggplots as variables (P1,P2,P#), which we then simply combine by P1+P2+P3 using the "patchwork" package.
By adding "coord_flip" to the ggplots, we swap the x and y axis. Then, by adding trans="reverse" we reverse the vertical axis, so that depth increases downwards.
```{r}
# vertical plotting
P1 <- ggplot(data = TEX, mapping = aes(x = `Depth mbsf`, y = `TEX`)) +
  geom_point(colour = "red") +
  geom_line(linetype = 2) +
  geom_ribbon(mapping = aes(ymin = (TEX - 0.06), ymax = (TEX + 0.06)), width = 10, alpha = 0.25, fill = "blue") +
  coord_flip(xlim = c(420, 80), ylim = c(0.2, 0.8)) +
  labs(x = "Depth (mbsf)", y = expression(TEX[86])) +
  scale_x_continuous(trans = "reverse", breaks = seq(420, 80, -25)) +
  theme_classic() +
  theme(
    axis.title = element_text(size = 10, colour = "black"),
    axis.text = element_text(size = 10, colour = "black"),
    axis.ticks = element_line(colour = "black")
  )

P2 <- ggplot(data = TEX) +
  geom_line(mapping = aes(x = `Depth mbsf`, y = `Methan index`), colour = "blue") +
  coord_flip(xlim = c(420, 80)) +
  labs(x = "Depth (mbsf)", y = "Methane Index") +
  scale_x_continuous(trans = "reverse", breaks = seq(420, 80, -25)) +
  theme_classic() +
  theme(
    axis.title.x = element_text(size = 10, colour = "black"),
    axis.text.x = element_text(size = 10, colour = "black"),
    axis.ticks.x = element_line(colour = "black"),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    axis.title.y = element_blank()
  )

P3 <- ggplot(data = tb_area, mapping = aes(x = depth, y = percent, fill = taxa)) +
  geom_area(position = "stack", orientation = "x") +
  geom_segment(alpha = 0.1, aes(y = 100, xend = depth, yend = 0)) +
  coord_flip(xlim = c(420, 80), ylim = c(0, 100)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(trans = "reverse", breaks = seq(420, 80, -25)) +
  labs(
    fill = "",
    y = "%",
    x = "Depth (mbsf)"
  ) +
  theme_classic() +
  scale_fill_brewer(palette = "Spectral") +
  theme(
    axis.title.x = element_text(size = 10, colour = "black"),
    axis.text.x = element_text(size = 10, colour = "black"),
    axis.ticks.x = element_line(colour = "black"),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    axis.line.y = element_blank(),
    legend.position = "right",
    legend.text = element_text(size = 8, colour = "black")
  )

P1 + P2 + P3 + plot_layout(ncol = 3)

ggsave("finalfigv.pdf")
```
